Reanalysis Step 4: Outcome Analysis with Three-Way Interaction

### 2025-08-19 reanalysis-2-multiple_imputation.qmd
style

# Print initial imputation model summary
message("Total variables in dataset: ", ncol(df_mice))
message("Variables to be imputed: ", sum(mice_dry$nmis > 0))
message("Total missing values: ", sum(mice_dry$nmis))
message("\nMissing data by variable:")
miss_vars <- data.frame(
  Variable = names(mice_dry$nmis),
  N_Missing = mice_dry$nmis,
  Pct_Missing = round(mice_dry$nmis/nrow(df_mice) * 100, 1)
) |>
  filter(N_Missing > 0) |>
  arrange(desc(N_Missing))

@@

message("\nDefault imputation methods:")
print(method_table)

@@

# Print final imputation model configuration
message("\n=== FINAL IMPUTATION MODEL ===\n")

@@

# Show which variables will be imputed
vars_to_impute <- names(custom_methods)[custom_methods != ""]
message("\nVariables that WILL be imputed (", length(vars_to_impute), " total):")
print(vars_to_impute)

# Show predictor matrix summary
message("\nPredictor matrix summary:")
message("- Each variable uses ", sum(pred_matrix[1,]), " predictors on average")
message("- Variables used as predictors:")
predictors <- names(which(colSums(pred_matrix) > 0))
print(predictors)

# Print model parameters
message("\n=== IMPUTATION PARAMETERS ===")
message("Mode: ", analysis_mode)
message("Number of imputations (m): ", n_imputations)
message("Max iterations per imputation (maxit): ", n_iterations)
message("Random seed: 2025")
```

@@

# Variables that were imputed
imputed_vars <- names(mice_results$nmis)[mice_results$nmis > 0]
message("\nVariables that were imputed:")
print(imputed_vars)
```

>>>>

# Print initial imputation model summary
message("Total variables in dataset: ", ncol(df_mice))
message("Variables to be imputed: ", sum(mice_dry$nmis > 0))
message("Total missing values: ", sum(mice_dry$nmis))
message("Missing data by variable:")
miss_vars <- data.frame(
  Variable = names(mice_dry$nmis),
  N_Missing = mice_dry$nmis,
  Pct_Missing = round(mice_dry$nmis/nrow(df_mice) * 100, 1)
) |>
  filter(N_Missing > 0) |>
  arrange(desc(N_Missing))

@@

message("Default imputation methods:")
print(method_table)

@@

# Print final imputation model configuration
message("FINAL IMPUTATION MODEL")

@@

# Show which variables will be imputed
vars_to_impute <- names(custom_methods)[custom_methods != ""]
message("Variables that WILL be imputed (", length(vars_to_impute), " total):")
print(vars_to_impute)

# Show predictor matrix summary
message("Predictor matrix summary:")
message("- Each variable uses ", sum(pred_matrix[1,]), " predictors on average")
message("- Variables used as predictors:")
predictors <- names(which(colSums(pred_matrix) > 0))
print(predictors)

# Print model parameters
message("=== IMPUTATION PARAMETERS ===")
message("Mode: ", analysis_mode)
message("Number of imputations (m): ", n_imputations)
message("Max iterations per imputation (maxit): ", n_iterations)
message("Random seed: 2025")
```

@@

# Variables that were imputed
imputed_vars <- names(mice_results$nmis)[mice_results$nmis > 0]
message("Variables that were imputed:")
print(imputed_vars)
```

### 2025-08-19 reanalysis-3-propensity_scores.qmd
style

# Run twang for each imputed dataset
message("\nCalculating propensity scores using Twang GBM...")
message("Parameters:")
message("  - n.trees: ", n_trees)
message("  - interaction.depth: 3")
message("  - shrinkage: 0.01")
message("  - estimand: ATT")
message("  - stop.method: es.mean (effect size only)")

for (i in 1:n_imputations) {
  message("\nProcessing imputation ", i, " of ", n_imputations, "...")

@@

message("\nTwang propensity score calculation complete")
```

@@

  # Print summary for diagnostic purposes (this will show balance)
  if (i <= max_imputations_for_diagnostics) {
    message("\nBalance summary for imputation ", i, ":")
    print(summary(twang_model))
  }
}

@@

message("\nWeight Summary Across All Imputations:")
kable(balance_summary_all, 
      caption = "Weight distribution by imputation",
      digits = 3)

@@

message("\nAverage Weights Across Imputations:")
kable(avg_weights,
      caption = "Average weight statistics across all imputations",
      digits = 3)

# Check weight stability
if (avg_weights$Max_Weight_Overall > 10) {
  message("\n⚠ Warning: Maximum weight > 10, indicating potential instability")
} else if (avg_weights$Max_Weight_Overall > 5) {
  message("\nNote: Maximum weight > 5, monitor for stability")
} else {
  message("\n✓ Weights appear stable (max < 5)")
}

message("\nNote: Balance statistics are shown in the summary output above.")
message("The twang package optimizes to minimize effect sizes (es.mean).")
```

@@

message("\nGenerating twang diagnostic plots for ", n_to_show, " imputations...")

# For each selected imputation, show twang-specific diagnostics
for (i in imps_to_show) {
  message("\nDiagnostics for Imputation ", i, ":")

@@

# Plot PS distribution for first few imputations
for (i in imps_to_show) {
  message("\nPropensity score distributions for imputation ", i)

@@

message("\nTwang propensity score results saved to: ", 
        file.path(reanalysis_data_dir, "ps_results_twang.rds"))
```

@@

```{r summary}
# Final summary
message("\n=== TWANG PROPENSITY SCORE ANALYSIS SUMMARY ===")
message("Method: Generalized Boosted Models (GBM) via twang")
message("Imputations processed: ", n_imputations)
message("Analysis mode: ", analysis_mode)
message("Trees used: ", n_trees)
message("Estimand: ATT (Average Treatment Effect on the Treated)")
message("Stop method: es.mean (effect size minimization)")

# Enhanced formula summary
message("\nEnhanced formula included:")
message("  - DR severity (person_dr)")
message("  - Treatment type")
message("  - All standard covariates from original analysis")

# Weight summary  
message("\nWeight Statistics (averaged across imputations):")
message("  - Mean weight (treated): ", round(avg_weights$Avg_Weight_Treated, 3))
message("  - Mean weight (control): ", round(avg_weights$Avg_Weight_Control, 3))
message("  - Maximum weight: ", round(avg_weights$Max_Weight_Overall, 2))

# Weight stability assessment
if (avg_weights$Max_Weight_Overall > 10) {
  message("\n⚠ Warning: Maximum weight > 10, indicating potential instability")
  message("  Consider increasing n.trees or adjusting twang parameters")
} else if (avg_weights$Max_Weight_Overall > 5) {
  message("\nNote: Maximum weight > 5, monitor for stability in outcome analysis")
} else {
  message("\n✓ Weights appear stable (max < 5)")
}

message("\nBalance Assessment:")
message("  See summary output above for detailed balance statistics.")
message("  The twang package optimizes es.mean to achieve covariate balance.")
```

>>>>

# Run twang for each imputed dataset
message("Calculating propensity scores using Twang GBM...")
message("Parameters:")
message("  - n.trees: ", n_trees)
message("  - interaction.depth: 3")
message("  - shrinkage: 0.01")
message("  - estimand: ATT")
message("  - stop.method: es.mean (effect size only)")

for (i in 1:n_imputations) {
  message("Processing imputation ", i, " of ", n_imputations, "...")

@@

message("Twang propensity score calculation complete")
```

@@

  # Print summary for diagnostic purposes (this will show balance)
  if (i <= max_imputations_for_diagnostics) {
    message("Balance summary for imputation ", i, ":")
    print(summary(twang_model))
  }
}

@@

message("Weight Summary Across All Imputations:")
kable(balance_summary_all, 
      caption = "Weight distribution by imputation",
      digits = 3)

@@

message("Average Weights Across Imputations:")
kable(avg_weights,
      caption = "Average weight statistics across all imputations",
      digits = 3)

# Check weight stability
if (avg_weights$Max_Weight_Overall > 10) {
  message("⚠ Warning: Maximum weight > 10, indicating potential instability")
} else if (avg_weights$Max_Weight_Overall > 5) {
  message("Note: Maximum weight > 5, monitor for stability")
} else {
  message("✓ Weights appear stable (max < 5)")
}

message("Note: Balance statistics are shown in the summary output above.")
message("The twang package optimizes to minimize effect sizes (es.mean).")
```

@@

message("Generating twang diagnostic plots for ", n_to_show, " imputations...")

# For each selected imputation, show twang-specific diagnostics
for (i in imps_to_show) {
  message("Diagnostics for Imputation ", i, ":")

@@

# Plot PS distribution for first few imputations
for (i in imps_to_show) {
  message("Propensity score distributions for imputation ", i)

@@

message("Twang propensity score results saved to: ", 
        file.path(reanalysis_data_dir, "ps_results_twang.rds"))
```

@@

```{r summary}
# Final summary
message("=== TWANG PROPENSITY SCORE ANALYSIS SUMMARY ===")
message("Method: Generalized Boosted Models (GBM) via twang")
message("Imputations processed: ", n_imputations)
message("Analysis mode: ", analysis_mode)
message("Trees used: ", n_trees)
message("Estimand: ATT (Average Treatment Effect on the Treated)")
message("Stop method: es.mean (effect size minimization)")

# Enhanced formula summary
message("Enhanced formula included:")
message("  - DR severity (person_dr)")
message("  - Treatment type")
message("  - All standard covariates from original analysis")

# Weight summary  
message("Weight Statistics (averaged across imputations):")
message("  - Mean weight (treated): ", round(avg_weights$Avg_Weight_Treated, 3))
message("  - Mean weight (control): ", round(avg_weights$Avg_Weight_Control, 3))
message("  - Maximum weight: ", round(avg_weights$Max_Weight_Overall, 2))

# Weight stability assessment
if (avg_weights$Max_Weight_Overall > 10) {
  message("⚠ Warning: Maximum weight > 10, indicating potential instability")
  message("  Consider increasing n.trees or adjusting twang parameters")
} else if (avg_weights$Max_Weight_Overall > 5) {
  message("Note: Maximum weight > 5, monitor for stability in outcome analysis")
} else {
  message("✓ Weights appear stable (max < 5)")
}

message("Balance Assessment:")
message("  See summary output above for detailed balance statistics.")
message("  The twang package optimizes es.mean to achieve covariate balance.")
```

### 2025-08-19 reanalysis-4-outcome_analysis.qmd
style

# Summary of first dataset for verification
first_data <- imputed_datasets[[1]]
message("\nFirst dataset summary:")
message("  - N observations: ", nrow(first_data))
message("  - N unique patients: ", length(unique(first_data$e_mrn_deidentified)))
message("  - Mean IPTW weight: ", round(mean(first_data$iptw_weight), 3))
message("  - Weight range: [", round(min(first_data$iptw_weight), 3), ", ", 
        round(max(first_data$iptw_weight), 3), "]")
```

@@

# Check distributions in first dataset
message("\nVariable distributions in first imputed dataset:")
message("\nDR Severity:")
print(table(imputed_datasets[[1]]$dr_severity))
message("\nTreatment Status:")
print(table(imputed_datasets[[1]]$any_treatment))
message("\nLapse Status:")
print(table(imputed_datasets[[1]]$ever_lapse_binary))

# Cross-tabulation of the three factors
message("\nThree-way cross-tabulation:")
with(imputed_datasets[[1]], 
     ftable(dr_severity, any_treatment, ever_lapse_binary))
```

@@

# Count interaction terms
message("\nInteraction structure:")
message("  - Main effects: 3 (lapse, DR severity [2 df], treatment)")
message("  - Two-way interactions: 3 (lapse×DR [2 df], lapse×treatment [1 df], DR×treatment [2 df])")
message("  - Three-way interaction: 1 (lapse×DR×treatment [2 df])")
message("  - Total interaction df: 7")
```

@@

message("\nFitting weighted models to each imputed dataset...")

for (i in 1:n_imputations) {
  message("\nImputation ", i, " of ", n_imputations, ":")

@@

message("\nAll models fitted successfully")
```

@@

# Display diagnostic information
message("\nPooling diagnostics:")
message("  - Number of imputations: ", n_imputations)
message("  - Number of parameters: ", nrow(pooled_results))
message("  - Average degrees of freedom: ", round(mean(pooled_results$df), 1))

@@

message("\nPerforming pooled joint hypothesis tests...")

@@

  message("\n1. Three-way interaction test (H0: all three-way terms = 0):")
  message("   Chi-sq = ", round(pooled_test_3way$D_bar, 3))
  message("   df = ", pooled_test_3way$df)
  message("   p-value = ", format.pval(pooled_test_3way$p_value, digits = 4, eps = 0.0001))

@@

  message("\n2. Lapse × DR severity interaction test:")
  message("   Chi-sq = ", round(pooled_test_lapse_dr$D_bar, 3))
  message("   df = ", pooled_test_lapse_dr$df)
  message("   p-value = ", format.pval(pooled_test_lapse_dr$p_value, digits = 4, eps = 0.0001))
}

@@

  message("\n3. Lapse × treatment interaction test:")
  message("   Chi-sq = ", round(pooled_test_lapse_tx$D_bar, 3))
  message("   df = ", pooled_test_lapse_tx$df)
  message("   p-value = ", format.pval(pooled_test_lapse_tx$p_value, digits = 4, eps = 0.0001))
}
```

@@

  message("\nCalculating marginal means and contrasts across all imputations...")
  message("This may take a few moments...")

@@

  message("\nCalculating Average Marginal Effects across all imputations...")

@@

```{r summary}
# Summarize key findings using POOLED results
message("\n=== OUTCOME ANALYSIS SUMMARY ===")
message("\nThree-Way Interaction Model Results:")
message("  - ", n_imputations, " imputed datasets analyzed")
message("  - IPTW weights from twang GBM applied")
message("  - Results properly pooled using Rubin's rules with Barnard-Rubin df adjustment")
message("  - Survey design accounts for patient clustering")

# Report pooling quality metrics
message("\nPooling Quality Metrics:")

@@

# Identify patterns from POOLED results
message("\nKey Patterns Observed (from pooled analysis):")

@@

message("\nStrongest lapse effect:")
strongest <- effect_summary[1, ]
message("  - ", strongest$DR_Severity, " + ", strongest$Treatment, 
        ": OR = ", strongest$OR, " (p ", strongest$P_Value, ")")

message("\nWeakest lapse effect:")
weakest <- effect_summary[nrow(effect_summary), ]
message("  - ", weakest$DR_Severity, " + ", weakest$Treatment, 
        ": OR = ", weakest$OR, " (p ", weakest$P_Value, ")")

# Clinical interpretation
message("\n=== CLINICAL INTERPRETATION ===")
message("\nThe three-way interaction reveals that:")
message("1. The harmful effect of lapsing varies by BOTH disease severity AND treatment status")
message("2. Patients with moderate disease (NPDR) show the largest vulnerability to lapsing")
message("3. Treatment appears to modify the lapse-outcome relationship differently across DR severity levels")
message("4. Advanced disease (PDR) patients show minimal harm from lapsing, regardless of treatment")

message("\nImplications for care continuity programs:")
message("- Resources should be prioritized for NPDR patients")
message("- Treatment status should inform retention strategies")
message("- PDR patients may have different care continuity needs")
```

@@

message("\nResults saved to: ", 
        file.path(reanalysis_data_dir, "outcome_analysis_results.rds"))
```

>>>>

# Summary of first dataset for verification
first_data <- imputed_datasets[[1]]
message("First dataset summary:")
message("  - N observations: ", nrow(first_data))
message("  - N unique patients: ", length(unique(first_data$e_mrn_deidentified)))
message("  - Mean IPTW weight: ", round(mean(first_data$iptw_weight), 3))
message("  - Weight range: [", round(min(first_data$iptw_weight), 3), ", ", 
        round(max(first_data$iptw_weight), 3), "]")
```

@@

# Check distributions in first dataset
message("Variable distributions in first imputed dataset:")
message("DR Severity:")
print(table(imputed_datasets[[1]]$dr_severity))
message("Treatment Status:")
print(table(imputed_datasets[[1]]$any_treatment))
message("Lapse Status:")
print(table(imputed_datasets[[1]]$ever_lapse_binary))

# Cross-tabulation of the three factors
message("Three-way cross-tabulation:")
with(imputed_datasets[[1]], 
     ftable(dr_severity, any_treatment, ever_lapse_binary))
```

@@

# Count interaction terms
message("Interaction structure:")
message("  - Main effects: 3 (lapse, DR severity [2 df], treatment)")
message("  - Two-way interactions: 3 (lapse×DR [2 df], lapse×treatment [1 df], DR×treatment [2 df])")
message("  - Three-way interaction: 1 (lapse×DR×treatment [2 df])")
message("  - Total interaction df: 7")
```

@@

message("Fitting weighted models to each imputed dataset...")

for (i in 1:n_imputations) {
  message("Imputation ", i, " of ", n_imputations, ":")

@@

message("All models fitted successfully")
```

@@

# Display diagnostic information
message("Pooling diagnostics:")
message("  - Number of imputations: ", n_imputations)
message("  - Number of parameters: ", nrow(pooled_results))
message("  - Average degrees of freedom: ", round(mean(pooled_results$df), 1))

@@

message("Performing pooled joint hypothesis tests...")

@@

  message("1. Three-way interaction test (H0: all three-way terms = 0):")
  message("   Chi-sq = ", round(pooled_test_3way$D_bar, 3))
  message("   df = ", pooled_test_3way$df)
  message("   p-value = ", format.pval(pooled_test_3way$p_value, digits = 4, eps = 0.0001))

@@

  message("2. Lapse × DR severity interaction test:")
  message("   Chi-sq = ", round(pooled_test_lapse_dr$D_bar, 3))
  message("   df = ", pooled_test_lapse_dr$df)
  message("   p-value = ", format.pval(pooled_test_lapse_dr$p_value, digits = 4, eps = 0.0001))
}

@@

  message("3. Lapse × treatment interaction test:")
  message("   Chi-sq = ", round(pooled_test_lapse_tx$D_bar, 3))
  message("   df = ", pooled_test_lapse_tx$df)
  message("   p-value = ", format.pval(pooled_test_lapse_tx$p_value, digits = 4, eps = 0.0001))
}
```

@@

  message("Calculating marginal means and contrasts across all imputations...")
  message("This may take a few moments...")

@@

  message("Calculating Average Marginal Effects across all imputations...")

@@

```{r summary}
# Summarize key findings using POOLED results
message("=== OUTCOME ANALYSIS SUMMARY ===")
message("Three-Way Interaction Model Results:")
message("  - ", n_imputations, " imputed datasets analyzed")
message("  - IPTW weights from twang GBM applied")
message("  - Results properly pooled using Rubin's rules with Barnard-Rubin df adjustment")
message("  - Survey design accounts for patient clustering")

# Report pooling quality metrics
message("Pooling Quality Metrics:")

@@

# Identify patterns from POOLED results
message("Key Patterns Observed (from pooled analysis):")

@@

message("Strongest lapse effect:")
strongest <- effect_summary[1, ]
message("  - ", strongest$DR_Severity, " + ", strongest$Treatment, 
        ": OR = ", strongest$OR, " (p ", strongest$P_Value, ")")

message("Weakest lapse effect:")
weakest <- effect_summary[nrow(effect_summary), ]
message("  - ", weakest$DR_Severity, " + ", weakest$Treatment, 
        ": OR = ", weakest$OR, " (p ", weakest$P_Value, ")")

# Clinical interpretation
message("=== CLINICAL INTERPRETATION ===")
message("The three-way interaction reveals that:")
message("1. The harmful effect of lapsing varies by BOTH disease severity AND treatment status")
message("2. Patients with moderate disease (NPDR) show the largest vulnerability to lapsing")
message("3. Treatment appears to modify the lapse-outcome relationship differently across DR severity levels")
message("4. Advanced disease (PDR) patients show minimal harm from lapsing, regardless of treatment")

message("Implications for care continuity programs:")
message("- Resources should be prioritized for NPDR patients")
message("- Treatment status should inform retention strategies")
message("- PDR patients may have different care continuity needs")
```

@@

message("Results saved to: ", 
        file.path(reanalysis_data_dir, "outcome_analysis_results.rds"))
```
